A thermodynamic view of dusty protoplanetary disks: axisymmetric stability

### 2016-08-26 model.tex
minor edits

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump, which then drags the 
gas towards it (and hence heating) to strengthen the pressure bump. A phase
difference was also noted in the numerical calculations of
\citet[][although they consider the full two-fluid system]{youdin07b}.

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, given the complexity of $\mathcal{C}$ (
Appendix \ref{lin_dust}), we might generally expect the system to
support a range of modes, stable and unstable. 
On physical grounds, growing modes  
\emph{must} be so by causing a phase lag between pressure and total
density.

>>>>

Indeed, \cite{jacquet11} explains the essence of the streaming 
instability in dusty protoplanetary disks as dust accumulation
(and hence compression) at a pressure bump, which then drags the 
gas towards it (and hence heating) to strengthen the pressure
bump. That is, heating occurs upon compression, as occurs for stellar 
pulsational instabilities. A phase difference was also noted in the 
numerical calculations of the streaming instability by 
\citet{youdin07b}.

We suggest understanding dusty instabilities in
protoplanetary disks by re-casting the problem into that of
thermodynamics. Then dust-drag instabilities appear in the equations in the 
same way as pulsational instabities in stars, which has already been 
well studied. Whether or not the dusty disk is actually unstable
depends on details of the dust-drag or `cooling' function
$\mathcal{C}$. However, given the complexity of $\mathcal{C}$ (see
Appendix \ref{lin_dust}), we might generally expect the system to 
support a range of stable and unstable modes. On physical grounds,
growing modes \emph{must} be so by causing a phase lag between
pressure and total density.

### 2016-08-26 summary.tex
minor edits

As noted in \S\ref{loc_iso_eos}, dust settling causes the effective
temperature to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
\citep[e.g.][]{mamat10, kim12,lin14c} where the temperature decreases
from the midplane. While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, having a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this linear problem in a following 
study.

@@

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas, and $\rhod/\rhog$ is conserved following the dusty flow. 
This is analogous to entropy being conserved following a adiabatic,
pure gas.

For $\tstop\neq 0$, the dust content of a
fluid parcel of the dusty mixture no longer conserved. The parcel is
is allowed to exchange dust particles with neighboring parcels owing
to finite dust-gas friction. This is analogous to heat exchange between a 
pure gas parcel and its surroundings. %Thus, finite dust-gas friction
%may be interepreted as an energy source for the dusty
%gas.

@@

We obtain the Solberg-Hoiland criteria to assess 
axisymmetric stability when the equation of state is strictly
isothermal. We show that {\bf the vertical shear associated with thin dust
 layers, however large it is, cannot lead to axisymmetric
  instabilities} \citep[cf. \emph{non-axisymmetric} Kelvin-Helmholtz instabilities
  in dusty disks, ][]{lee10}. We show that
sharp edges in the dust-to-gas ratio cannot persist in protoplanetary
disks, as these imply a sharp entropy gradient and may thus destablize
the disk.

@@

For locally isothermal disks $c^2_s=c_s^2(r)$ we generalize the vertical
shear instability (VSI) to include dust. We find that dust-loading
generally has a stabilizing effect. In our disk models
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}.

@@

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. In particular, we emphasize  that the 
physical origin for any instability associated with strong dust-gas
drag is a phase lag between the pressure and total density evolution
of fluid elements. This leads to growing oscillations as positive work
is done, provided by dust-gas drag, which is equivalent to a `heat'
source. %We demonstrate a potentially new dust-drag instability in
%stratified disks that may explain

>>>>

As noted in \S\ref{loc_iso_eos}, dust settling causes the effective
temperature to \emph{increase} away from the midplane. This contrasts
to previous studies of GI in vertically stratified disks
 where the temperature decreases
from the midplane \citep[e.g.][]{mamat10, kim12,lin14c}. 
While we expect only the total surface density and
characteristic temperatures are relevant to stability  
\citep{toomre64}, having a non-trivial vertical temperature
structure, induced by dust, may modify the vertical structure of 3D
waves and unstable modes. We will examine this linear problem in a following 
study.

@@

\section{Summary}\label{summary}
In this paper we highlight the similarity between an isothermal 
dusty gas and an adiabatic, pure gas. The correspondence arises
because drag forces reduce the relative velocity between gas and
dust. In the limit of perfect dust-gas coupling, $\tstop \to 0$,  
 dust is entrained in 
the gas, and the dust-to-gas ratio $\rhod/\rhog$ is conserved
following the dusty flow. This is analogous to entropy being conserved
following a adiabatic, pure gas.

For finite dust-gas friction, $\tstop\neq 0$, the dust content of a
fluid parcel of the dusty mixture no longer conserved. A fluid parcel 
can exchange dust particles with neighboring parcels owing 
to dust-gas relative drift. This is analogous to heat exchange between
a  pure gas parcel and its surroundings.

@@

We obtain the Solberg-Hoiland criteria for axisymmetric stability for
strictly isothermal, perfectly-coupled dusty gas.  
We show that {\bf the vertical shear associated with thin dust
 layers, however large it is, cannot lead to axisymmetric
  instabilities}. This is consistent with the fact that only 
\emph{non-axisymmetric} Kelvin-Helmholtz instabilities have been 
observed in numerical simulations in the perfectly-coupled limit. 
\citep{chiang08,barranco09,lee10}. We show that 
sharp radial edges in the dust-to-gas ratio cannot persist in
protoplanetary disks, as these imply a sharp entropy gradient and may
thus destablize the disk.

@@

For locally isothermal disks, $c^2_s=c_s^2(r)$, we generalize the 
vertical shear instability (VSI) to dusty disks, which is still 
driven by the radial temperature gradient. We find that 
dust-loading is generally stabilizing. In our disk models 
dust-loading does not affect VSI growth rates significantly, but
meridional motions may be suppressed at heights where dust-induced
vertical buoyancy dominates over vertical shear, consistent with our
previous study \citep{lin15}.

@@

We also discuss future applications of our thermodynamic framework to 
study dusty protoplanetary disks. In particular, we emphasize the 
physical origin of any instability driven by strong dust-gas
drag is a phase lag between the pressure and total density evolution
of the dusty gas mixture. This leads to growing modes because 
positive work is done during an oscillation cycle, provided by
dust-gas drag, which is equivalent to a `heat' 
source. This interpretation is entirely analogous to pulsational
instabilities in stars.

%We demonstrate a potentially new dust-drag instability in
%stratified disks that may explain
